

# Ask  

***

### About  

[Bellabeat](https://bellabeat.com/) is the go-to wellness brand for women with 
an ecosystem of products and services that focuses on women’s health. They 
empower women with knowledge about their own health and habits by collecting 
data on activity, sleep, stress, and reproductive health. Urška Sršen, co-founder 
and Chief Creative Officer of Bellabeat, believes that analyzing smart device 
fitness data could help unlock new growth opportunities for the company.  


### Key stakeholders  

The key stakeholders for this project include:  

* **Urška Sršen**: Bellabeat’s co-founder and Chief Creative Officer  
* **Sando Mur**: Mathematician and Bellabeat’s co-founder; key member of the 
Bellabeat executive team  
* **Bellabeat marketing analytics team**: A team of data analysts responsible 
for collecting, analyzing, and reporting data that helps guide Bellabeat’s 
marketing strategy  



### Guiding questions  

These three questions will guide our analysis:  

* What are some trends in smart device usage?  
* How could these trends apply to Bellabeat customers?  
* How could these trends help influence Bellabeat marketing strategy?  


### Business task  

>**Analyze smart device fitness data to gain insight into consumers and identify 
trends that can inform Bellabeat App marketing strategy.**  



# Prepare  

***



### Why R?  

We are using R for our data analysis because it’s a powerful, all-in-one tool. 
We can explore all the datasets quickly, provide code, share visualizations, 
and keep everything organized all within this html file. This html file will 
serve as a record of the data cleaning process, data analysis and data 
visualization generation that preceded our stakeholder presentation.  



### Install and load packages  

The first thing we'll do is install the needed packages:  

* **Install R packages**  
    + `install.packages()`  
    + Packages: tidyverse, ggplot2, here , hms, janitor, lubridate, skimr, \
    gridExtra, paletteer, crayon, stringr, patchwork, ggthemes, ggforce, \
    personograph, ggbeeswarm, knitr  

```{r, install-packages, eval=FALSE, include=FALSE}
# Install tidyverse package ---------------------------
# Includes ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats
install.packages("tidyverse") 
# Install additional packages ---------------------------
install.packages(c(
  "here", "hms", "janitor", "lubridate", "skimr", "gridExtra", "paletteer",
  "crayon", "stringr", "personograph", "patchwork", "ggthemes", "ggforce",
  "ggbeeswarm", "knitr",
))
```
  
Then we'll load the libraries for these packages:  

* **Load R libraries**  
    + `library()`  
    + Packages: tidyverse, ggplot2, here , hms, janitor, lubridate, skimr, \
    gridExtra, paletteer, crayon, stringr, patchwork, ggthemes, ggforce, \
    personograph, ggbeeswarm, knitr   

```{r, load-libraries, message=FALSE, include=FALSE}
# Load tidyverse ---------------------------
library("tidyverse")
# Load additional packages ---------------------------
library("here")
library("hms")
library("janitor")
library("lubridate")
library("skimr")
library("gridExtra")
library("paletteer")
library("crayon")
library("stringr")
library("patchwork")
library("ggthemes")
library("ggforce")
library("ggbeeswarm")
library("personograph")
library("knitr")
```
  


### Data sources  

We'll be using public data that explores smart device users' daily habits:  

* [**FitBit Fitness Tracker Data**](https://www.kaggle.com/arashnic/fitbit) (CC0: 
Public Domain, dataset made available through [Mobius](https://www.kaggle.com/arashnic)): 
This Kaggle data set contains personal tracker data from thirty Fitbit users. 
These Fitbit users consented to submission of personal tracker data, including 
minute-level output for physical activity, heart rate, and sleep monitoring.  
  





### Importing data  

Back to the FitBit data: we're going to explore data across fourteen CSV files. 
The data is organized in rows and columns, making it structured data. 
The majority of datasets are in long format, meaning each row is one time point 
per session ID. Each session ID has data in multiple rows. Individual reports 
can be parsed by export ID or time-stamp. These datasets fall into three 
different groups based on the time increments they were tracked by:  

1. Daily data  
2. Hourly data  
3. Minute data  


By analyzing the daily, hourly, and minute level data provided, we can begin to 
explore trends and search for insights that can be applied to Bellabeat 
customers.  
 

Let's look at the file sizes prior to importing. We don't want to overload our 
RAM in Rstudio Cloud:  


```{r, preview-file-size, echo=FALSE}
# Daily data
cat("dailyActivity_merged.csv", "=", sep = " ", file.info("fitbit_data/dailyActivity_merged.csv")$size, "bytes", fill = TRUE)
cat("dailyCalories_merged.csv", "=", sep = " ", file.info("fitbit_data/dailyCalories_merged.csv")$size, "bytes", fill = TRUE)
cat("dailyIntensities_merged.csv", "=", sep = " ", file.info("fitbit_data/dailyIntensities_merged.csv")$size, "bytes", fill = TRUE)
cat("dailySteps_merged.csv", "=", sep = " ", file.info("fitbit_data/dailySteps_merged.csv")$size, "bytes", fill = TRUE)
cat("sleepDay_merged.csv", "=", sep = " ", file.info("fitbit_data/sleepDay_merged.csv")$size, "bytes", fill = TRUE)
cat("weightLogInfo_merged.csv", "=", sep = " ", file.info("fitbit_data/weightLogInfo_merged.csv")$size, "bytes", fill = TRUE)
# Hourly data
cat("hourlyCalories_merged.csv", "=", sep = " ", file.info("fitbit_data/hourlyCalories_merged.csv")$size, "bytes", fill = TRUE)
cat("hourlyIntensities_merged.csv", "=", sep = " ", file.info("fitbit_data/hourlyIntensities_merged.csv")$size, "bytes", fill = TRUE)
cat("hourlySteps_merged.csv", "=", sep = " ", file.info("fitbit_data/hourlySteps_merged.csv")$size, "bytes", fill = TRUE)
# Minute data
cat("minuteCaloriesNarrow_merged.csv", "=", sep = " ", file.info("fitbit_data/minuteCaloriesNarrow_merged.csv")$size, "bytes", fill = TRUE)
cat("minuteIntensitiesNarrow_merged.csv", "=", sep = " ", file.info("fitbit_data/minuteIntensitiesNarrow_merged.csv")$size, "bytes", fill = TRUE)
cat("minuteMETsNarrow_merged.csv", "=", sep = " ", file.info("fitbit_data/minuteMETsNarrow_merged.csv")$size, "bytes", fill = TRUE)
cat("minuteSleep_merged.csv", "=", sep = " ", file.info("fitbit_data/minuteSleep_merged.csv")$size, "bytes", fill = TRUE)
cat("minuteStepsNarrow_merged.csv", "=", sep = " ", file.info("fitbit_data/minuteStepsNarrow_merged.csv")$size, "bytes", fill = TRUE)
```

Based on file sizes, it looks like the minute level data could cause issues with 
RAM usage in RStudio Cloud. Let's compare the size of the smallest minute level 
dataset to the total combined size of all nine daily/hourly datasets:  



* `minuteSleep_merged.csv` = **`8,848,370 bytes`**  


* `dailyActivity_merged.csv`, \
`dailyCalories_merged.csv`, \
`dailyIntensities_merged.csv`, \
`dailySteps_merged.csv`, \
`sleepDay_merged.csv`, \
`weightLogInfo_merged.csv`, \
`hourlyCalories_merged.csv`, \
`hourlyIntensities_merged.csv`, \
`hourlySteps_merged.csv` =  **`2,753,820 bytes`**  (`r sum(111288+25125+70581+25175+18100+6725+801486+898778+796562)`)
  


>**Even if we combine the bytes for all the daily and hourly datasets, it still doesn't even come close in size to the smallest minute level dataset!**   



Let's start with loading our daily and hourly datasets. We'll create dataframes 
(df) for these now:  

```{r, load-datasets-pt-1, message=FALSE, results="hide", eval=TRUE}
daily_activity <- read_csv("fitbit_data/dailyActivity_merged.csv")
daily_calories <- read_csv("fitbit_data/dailyCalories_merged.csv")
daily_intensities <- read_csv("fitbit_data/dailyIntensities_merged.csv")
daily_steps <- read_csv("fitbit_data/dailySteps_merged.csv")
daily_sleep <- read_csv("fitbit_data/sleepDay_merged.csv")
weight_log <- read_csv("fitbit_data/weightLogInfo_merged.csv")
hourly_calories <- read_csv("fitbit_data/hourlyCalories_merged.csv")
hourly_intensities <- read_csv("fitbit_data/hourlyIntensities_merged.csv")
hourly_steps <- read_csv("fitbit_data/hourlySteps_merged.csv")
```
<br>

Since the minute data is huge, we're only going to import the smallest dataset:  

```{r, load-datasets-pt-2, message=FALSE, results="hide", eval=TRUE}
minute_sleep <- read_csv("fitbit_data/minuteSleep_merged.csv")
```



# Process  

***



### Cleaning data  

Now that we've identified and loaded the ten datasets we're going to use, we 
need to ensure that the data within each  dataframe is clean and ready to 
analyze.  



#### Create tibbles  {.tabset .tabset-pills}

Let's take a closer look at the different dataframes. We'll start by using the 
`head()` function to view the first few rows of each one:  
 

##### Daily dataframes  



    `head(daily_activity)`  
    ```{r, preview-daily-activity, echo=FALSE, eval=TRUE}
    head(daily_activity)
    ```
    
    `head(daily_calories)`  
    ```{r, preview-daily-calories, echo=FALSE, eval=TRUE}
    head(daily_calories)
    ```
    
    `head(daily_intensities)`  
    ```{r, preview-daily-intensities, echo=FALSE, eval=TRUE}
    head(daily_intensities)
    ```
    
    `head(daily_steps)`  
    ```{r, preview-daily-steps, echo=FALSE, eval=TRUE}
    head(daily_steps)
    ```
    
    `head(daily_sleep)`  
    ```{r, preview-sleep-day, echo=FALSE, eval=TRUE}
    head(daily_sleep)
    ```
    
    `head(weight_log)`  
    ```{r, preview-weight-log, echo=FALSE, eval=TRUE}
    head(weight_log)
    ```



##### Hourly dataframes  



    `head(hourly_calories)`  
    ```{r, preview-hourly-calories, echo=FALSE, eval=TRUE}
    head(hourly_calories)
    ```
    
    `head(hourly_intensities`  
    ```{r, preview-hourly-intensities, echo=FALSE, eval=TRUE}
    head(hourly_intensities)
    ```
    
    `head(hourly_steps)`  
    ```{r, preview-hourly-steps, echo=FALSE, eval=TRUE}
    head(hourly_steps)
    ```

 

##### Minute dataframe  



    `head(minute_sleep)`  
    ```{r, preview-minute-sleep, echo=FALSE, eval=TRUE}
    head(minute_sleep)
    ```



#### View column names  {.tabset .tabset-pills}  

Now we'll use the `colnames()` function to view the column names for each 
dataframe:  


##### Daily dataframes  



    `colnames(daily_activity)`
    ```{r, daily-activity-colnames, echo=FALSE, eval=TRUE}
      colnames(daily_activity)
    ```
    
    `colnames(daily_calories)`  
    ```{r, daily-calories-colnames, echo=FALSE, eval=TRUE}
    colnames(daily_calories)
    ```
    
    `colnames(daily_intensities)`
    ```{r, daily-intensities-colnames, echo=FALSE, eval=TRUE}
    colnames(daily_intensities)
    ```
    
    `colnames(daily_steps)`
    ```{r, daily-steps-colnames, echo=FALSE, eval=TRUE}
    colnames(daily_steps)
    ```
    
    `colnames(daily_sleep)`
    ```{r, sleep-day-colnames, echo=FALSE, eval=TRUE}
    colnames(daily_sleep)
    ```
    
    `colnames(weight_log)`
    ```{r, weight-log-colnames, echo=FALSE, eval=TRUE}
    colnames(weight_log)
    ```

  

##### Hourly dataframes    



    `colnames(hourly_calories)`  
    ```{r, hourly-calories-colnames, echo=FALSE, eval=TRUE}
    colnames(hourly_calories)
    ```
    
    `colnames(hourly_intensities)`   
    ```{r, hourly-intensities-colnames, echo=FALSE, eval=TRUE}
    colnames(hourly_intensities)
    ```
    
    `colnames(hourly_steps)`  
    ```{r, hourly-steps-colnames, echo=FALSE, eval=TRUE}
    colnames(hourly_steps)
    ```



##### Minute dataframe  



    `colnames(minute_sleep)`
    ```{r, minute-sleep-colnames, echo=FALSE, eval=TRUE}
    colnames(minute_sleep)
    ```

   

#### View variables  {.tabset .tabset-pills}

All these dataframes have the `Id` field, which can be used to merge them into 
one another to create overall dataframes (i.e. daily df, hourly df). However, 
the activity date/time variable formatting looks to be different across some of 
the dataframes. Let's `select()` those variables to take a closer look.  

  
##### Daily dataframes  



    ```{r, select-daily-datasets, eval=TRUE}
daily_activity %>%
  select(Id, ActivityDate)
daily_calories %>%
  select(Id, ActivityDay)
daily_intensities %>%
  select(Id, ActivityDay)
daily_steps %>%
  select(Id, ActivityDay)
daily_sleep %>%
  select(Id, SleepDay)
weight_log %>%
  select(Id, Date)
    ```



##### Hourly dataframes  



    ```{r, select-hourly-datasets, eval=TRUE}
hourly_calories %>%
  select(Id, ActivityHour)
hourly_intensities %>%
  select(Id, ActivityHour)
hourly_steps %>%
  select(Id, ActivityHour)
    ```


##### Minute dataframe  



    ```{r, select-minute-datasets, eval=TRUE}
minute_sleep %>%
  select(Id, date)
    ```

#### Data cleaning {.tabset .tabset-pills}

Now that we've familiarized ourselves with the data, we can identify a few data
cleaning tasks to move forward with. Here are the main areas we'll address:  

1. **Naming conventions:** we'll adjust the variable names according to the 
[tidyverse style guide](https://style.tidyverse.org/syntax.html). All variable 
names will use only lowercase letters and underscores `_` (known as snake case) 
to separate words within a name. This will make our code easier to read.  

2. **Fix formatting issues:** we'll make the date-time format consistent across 
each dataframe. This will prevent problems down the line during analysis, as 
well as when we begin merging the data together.  

3. **Create new variables:** we'll create a few new columns within each 
dataframe.  

4. **Create new dataframe:** in order to better analyze the `minute_sleep` data, 
we'll first do some thorough transformation by reformatting and creating new 
variables. Then we'll aggregate all the sleep data we have into a more 
manageable structure.  
<br>

##### Daily dataframes  

```{r, format-daily-data, message=FALSE, eval=TRUE}
# daily_activity ---------------------------
daily_activity <-
  daily_activity %>% 
  rename(
    activity_date = ActivityDate, 
    total_steps = TotalSteps, 
    total_distance = TotalDistance,
    tracker_distance = TrackerDistance,
    logged_activities_d = LoggedActivitiesDistance,
    very_active_d = VeryActiveDistance, 
    moderately_active_d = ModeratelyActiveDistance, 
    light_active_d = LightActiveDistance, 
    sedentary_d = SedentaryActiveDistance, 
    very_active_m = VeryActiveMinutes, 
    fairly_active_m = FairlyActiveMinutes, 
    lightly_active_m = LightlyActiveMinutes, 
    sedentary_m = SedentaryMinutes, 
    calories = Calories
    ) %>% 
  rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    # reformat variable as POSIXct to represent date and time
    activity_date = parse_date_time(activity_date, "%m/%d/%y"),
    # create new variable and format as date only
    activity_date_ymd = as.Date(activity_date, "%Y/%m/%d"),
    # create new variables for day of week and time of week
    day_of_week = weekdays(as.Date(activity_date)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend")
    )
# daily_calories ---------------------------
daily_calories <-
  daily_calories %>% 
  rename(
    activity_date = ActivityDay,
    calories = Calories
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_date = parse_date_time(activity_date, "%m/%d/%Y"),
    activity_date_ymd = as.Date(activity_date, "%Y/%m/%d"),
    day_of_week = weekdays(as.Date(activity_date)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend")
  )
# daily_intensities ---------------------------
daily_intensities <-
  daily_intensities %>% 
  rename(
    activity_date = ActivityDay,
    very_active_d = VeryActiveDistance, 
    moderately_active_d = ModeratelyActiveDistance, 
    light_active_d = LightActiveDistance, 
    sedentary_d = SedentaryActiveDistance, 
    very_active_m = VeryActiveMinutes, 
    fairly_active_m = FairlyActiveMinutes, 
    lightly_active_m = LightlyActiveMinutes, 
    sedentary_m = SedentaryMinutes, 
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_date = parse_date_time(activity_date, "%m/%d/%Y"),
    activity_date_ymd = as.Date(activity_date, "%Y/%m/%d"),
    day_of_week = weekdays(as.Date(activity_date)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend")
  )
# daily_sleep ---------------------------
daily_sleep <-
  daily_sleep %>% 
  rename(
    activity_date = SleepDay,
    total_sleep_records = TotalSleepRecords,
    total_minutes_asleep = TotalMinutesAsleep,
    total_time_in_bed = TotalTimeInBed
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_date = parse_date_time(activity_date, "%m/%d/%Y %I:%M:%S %p"),
    activity_date_ymd = as.Date(activity_date, "%Y/%m/%d"),
    day_of_week = weekdays(as.Date(activity_date)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend")
    )
# daily_steps ---------------------------
daily_steps <-
  daily_steps %>% 
  rename(
    activity_date = ActivityDay,
    step_total = StepTotal
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_date = parse_date_time(activity_date, "%m/%d/%Y"),
    activity_date_ymd = as.Date(activity_date, "%Y/%m/%d"),
    day_of_week = weekdays(as.Date(activity_date)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend")
  )
# weight_log ---------------------------
weight_log <-
  weight_log %>% 
  rename(
    activity_date = Date,
    weight_kg = WeightKg,
    weight_lb = WeightPounds,
    fat = Fat,
    bmi = BMI,
    manual_report = IsManualReport,
    log_id = LogId
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_date = parse_date_time(activity_date, "%m/%d/%Y %I:%M:%S %p"),
    activity_date_ymd = as.Date(activity_date, "%Y/%m/%d"),
    activity_time = format(activity_date, format = "%I:%M:%S %p"), 
    day_of_week = weekdays(as.Date(activity_date)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend"),
    hour_of_day = as.POSIXct(activity_date, format = "%I:%M:%S %p")
  )
```



##### Hourly dataframes  

```{r, format-hourly-data, message=FALSE, eval=TRUE}
# hourly_calories ---------------------------
hourly_calories <-
  hourly_calories %>% 
  rename(
    activity_hour = ActivityHour,
    calories = Calories
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_hour = parse_date_time(activity_hour, "%m/%d/%Y %I:%M:%S %p"),
    activity_date_ymd = as.Date(activity_hour, "%Y/%m/%d"), 
    activity_time = format(activity_hour, format = "%I:%M:%S %p"),
    day_of_week = weekdays(as.Date(activity_hour)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend"),
    hour_of_day = as.POSIXct(activity_time, format = "%I:%M:%S %p")
  )
# hourly_intensities ---------------------------
hourly_intensities <- 
  hourly_intensities %>% 
  rename(
    activity_hour = ActivityHour,
    total_intensity = TotalIntensity,
    average_intensity = AverageIntensity
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_hour = parse_date_time(activity_hour, "%m/%d/%Y %I:%M:%S %p"),
    activity_date_ymd = as.Date(activity_hour, "%Y/%m/%d"), 
    activity_time = format(activity_hour, format = "%I:%M:%S %p"),
    day_of_week = weekdays(as.Date(activity_hour)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend"),
    hour_of_day = as.POSIXct(activity_time, format = "%I:%M:%S %p")
  )
# hourly_steps --------------------------- 
hourly_steps <- 
  hourly_steps %>% 
  rename(
    activity_hour = ActivityHour,
    step_total = StepTotal
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_hour = parse_date_time(activity_hour, "%m/%d/%Y %I:%M:%S %p"),
    activity_date_ymd = as.Date(activity_hour, "%Y/%m/%d"), 
    activity_time = format(activity_hour, format = "%I:%M:%S %p"),
    day_of_week = weekdays(as.Date(activity_hour)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend"),
    hour_of_day = as.POSIXct(activity_time, format = "%I:%M:%S %p")
  )
```



##### Minute dataframes  

```{r, format-minute-data, message=FALSE, eval=TRUE}
# minute_sleep ---------------------------
minute_sleep <-
  minute_sleep %>% 
  rename(
    activity_date = date,
    sleep_value = value,
    log_id = logId
  ) %>% 
   rename_with(
    tolower, starts_with("Id")
  ) %>% 
  mutate(
    activity_date = parse_date_time(activity_date, "%m/%d/%Y %I:%M:%S %p"),
    activity_date_ymd = as.Date(activity_date, "%Y/%m/%d"),
    activity_time = format(activity_date, format = "%I:%M:00 %p"), 
    day_of_week = weekdays(as.Date(activity_date)),
    time_of_week = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend"),
    hour_of_day = as.POSIXct(activity_time, format = "%I:%M:%S %p"),
    # sleep_id will be used generate totals per sleep log
    sleep_id = str_c(id, "-", log_id), 
    # Create new variables for sleep values - asleep, restless, and awake
    asleep = ifelse(sleep_value == 1, 1, 0),
    restless = ifelse(sleep_value == 2, 1, 0),
    awake = ifelse(sleep_value == 3, 1, 0)
  )
# Create sleep_summary_0 df ---------------------------
# that shows totals for 3 sleep values per sleep log
sleep_summary_0 <-
  minute_sleep %>%
  # id_date will be used to generate a daily total
  mutate(
    id_date = str_c(id, "-", activity_date_ymd)
  ) %>% 
  group_by(sleep_id, activity_date_ymd, id_date, id) %>% 
  dplyr::summarize(
    total_asleep = sum(sleep_value == "1"),
    total_restless = sum(sleep_value == "2"),
    total_awake = sum(sleep_value == "3")
  )
# Create sleep_summary df ---------------------------
# that combines totals for each day per id
sleep_summary <-
  sleep_summary_0 %>%
  # activity_date will be used to merge with daily_sleep df
  mutate(
    activity_date = parse_date_time(activity_date_ymd, "%Y/%m/%d")
  ) %>%
  group_by(id_date, activity_date, id) %>%
  dplyr::summarize(
    total_asleep_merged = sum(total_asleep),
    total_restless_merged = sum(total_restless),
    total_awake_merged = sum(total_awake)
  )
# Merge these two daily sleep dfs into one ---------------------------
sleep_data <- merge(x = daily_sleep, y = sleep_summary, by = c("id", "activity_date"), all = TRUE)
```




# Analyze  

***


## Summary statistics 

Now that we've done some deeper cleaning and formatting, let's get to know this 
data from a big-picture perspective. We can do this by obtaining some high-level 
summary statistics that can show us how these dataframes relate to one another, 
as well as what trends might exist within each:  



### A. Unique participants {.tabset .tabset-pills}  

How many unique participants are there in each dataframe?  

  
#### Daily dataframes  

```{r, distinct-daily-data, eval=TRUE}
# There are 33 users (one user per unique id) in the daily activity df
n_distinct(daily_activity$id)
# There are 24 users (one user per unique id) in the sleep dfs
n_distinct(daily_sleep$id)
n_distinct(sleep_data$id)
# There are 8 users (one user per unique id) in the weight df
n_distinct(weight_log$id)
```



#### Hourly dataframes    

```{r, distinct-hourly-data, eval=TRUE}
# There are 33 users (one user per unique id) in the hourly dfs
n_distinct(hourly_calories$id)
n_distinct(hourly_intensities$id)
n_distinct(hourly_steps$id)
```



#### Minute dataframe  

```{r, distinct-minute-data, eval=TRUE}
# There are 24 users (one user per unique id) in the minute df
n_distinct(minute_sleep$id)
```
  


### B. Total observations {.tabset .tabset-pills}  

How many observations are there in each dataframe?  

  
#### Daily dataframes  

```{r, observations-daily-data, eval=TRUE}
# The observations vary across the daily dfs
nrow(daily_activity)
nrow(daily_sleep)
nrow(sleep_data)
nrow(weight_log)
```



#### Hourly dataframes  

```{r, observations-hourly-data, eval=TRUE}
# There are 22099 observations in each hourly dataframe
nrow(hourly_calories)
nrow(hourly_intensities)
nrow(hourly_steps)
```



#### Minute dataframe  

```{r, observations-minute-data, eval=TRUE}
# There are 188521 observations in the minute dataframe
nrow(minute_sleep)
```
   

 
### C. Quick overview {.tabset .tabset-pills}  

What are some quick summary statistics we'd want to know about each dataframe?  

#### Daily dataframes  

```{r, summary-stats-daily-data, message=FALSE, eval=TRUE}
# Daily totals for steps, distance, calories ---------------------------
daily_activity %>%
  select(
    total_steps,
    total_distance,
    calories
  ) %>%
  summary()
# Active minute levels per category ---------------------------
daily_activity %>%
  select(
    very_active_m,
    fairly_active_m,
    lightly_active_m,
    sedentary_m
  ) %>%
  summary()
# Sleep totals ---------------------------
# for records, minutes asleep, and time in bed
daily_sleep %>%
  select(
    total_sleep_records,
    total_minutes_asleep,
    total_time_in_bed
  ) %>%
  summary()
# Sleep totals per category ---------------------------
sleep_data %>%
  select(
    total_minutes_asleep, 
    total_time_in_bed,
    total_asleep_merged,
    total_restless_merged,
    total_awake_merged
  ) %>%
  summary()
# Weight totals ---------------------------
weight_log %>%
  select(
    weight_lb,
    fat,
    bmi
  ) %>%
  summary()
```



#### Hourly dataframes  

```{r, summary-stats-hourly-data, message=FALSE, eval=TRUE}
# Hourly calories summary ---------------------------
hourly_calories %>%
  select(calories) %>%
  summary()
# Hourly intensities summary ---------------------------
hourly_intensities %>%
  select(
    total_intensity,
    average_intensity
  ) %>%
  summary()
# Hourly steps summary ---------------------------
hourly_steps %>%
  select(step_total) %>%
  summary()
```



#### Minute dataframe  

```{r, summary-stats-minute-data, message=FALSE, eval=TRUE}
# Minute sleep summary ---------------------------
minute_sleep %>%
  select(
    sleep_value,
    asleep,
    restless,
    awake
  ) %>%
  summary()
```
  


### D. Interesting notes  

What does this tell us about this sample of people's activities? A few 
interesting notes:  
  
* **7,638** - average total steps per day, which is below the generally 
recommended 10,000 goal.  
* **991.2** - average "Sedentary" minutes per day, which is the highest average 
of the tracked activity minutes per day. This is just over 16.5 hours per day.  
* **192.8**- average "Lightly Active" minutes per day, which the highest among 
the non-sedentary activity minutes per day (*21.16* - Very Active, *13.56*- 
Fairly Active).  
* **34.72** - combined daily average of Very Active (21.16) and Fairly Active 
(13.56) minutes per day. This indicates that, on average, these individuals may 
mostly be getting the WHO recommended 150 minutes of moderate-intensity physical 
activity per week [Global recommendations on physical activity for health](https://www.who.int/publications/i/item/9789241599979)  
* **91.5%** - average percentage of time asleep while in bed. This is calculated 
by dividing "Total Minutes Asleep" (419.5) by "Total Time In Bed" (458.6).  
* **97.39** - average calories burned per hour.  
  

  
## Plotting a few explorations  



### Steps and active minutes  

The goal of 10,000 steps per day has been widely proposed across the smart 
device industry. But how important of a health metric is daily steps really? 
For example, what's the relationship between total steps and calories burned?  

```{r, total-steps-vs-calories, message=FALSE, eval=TRUE}
# Total steps vs calories plot ---------------------------
ggplot(data = daily_activity, aes(x = total_steps, y = calories)) +
  geom_point() +
  geom_smooth() + # Trend line with a shadow representing  95% confidence interval
  labs(title = "Total steps vs calories") +
  ylab("Calories") +
  xlab("Total Steps") +
  theme_minimal()
cor.test(daily_activity$total_steps, daily_activity$calories, method = "pearson")
```

As expected, we see a positive correlation here between `total_steps` and 
`calories`. This makes sense because the more active someone is, the more 
calories they will burn. But what about the relationship between 
`very_active_m` and `calories`?  

```{r, very-active-minutes-vs-calories, eval=TRUE}
# Very active minutes vs calories plot ---------------------------
ggplot(data = daily_activity, aes(x = very_active_m, y = calories)) +
  geom_point() +
  geom_smooth() + # Trend line with a shadow representing  95% confidence interval
  labs(title = "Very active minutes vs calories") +
  ylab("Calories") +
  xlab("Very active minutes") +
  theme_minimal()
cor.test(daily_activity$very_active_m, daily_activity$calories, method = "pearson")
```

Unsurprisingly, we see another positive correlation. This time between 
`very_active_m` and `calories`. Very active people will burn more 
calories. But we know from earlier that the average `very_active_m` per day 
of `21.16` is much lower than then average `lightly_active_m` per day of 
`192.80`. So what's the relationship between `lightly_active_m` and 
`calories`?  

```{r, lightly-active-minutes-vs-calories, eval=TRUE}
# Lightly active minutes vs calories plot ---------------------------
ggplot(data = daily_activity, aes(x = lightly_active_m, y = calories)) +
  geom_point() +
  geom_smooth() + # Trend line with a shadow representing  95% confidence interval
  labs(title = "Lightly active minutes vs calories") +
  ylab("Calories") +
  xlab("Lightly active minutes") +
  theme_minimal()
cor.test(daily_activity$lightly_active_m, daily_activity$calories, method = "pearson")
```

The relationship between `lightly_active_m` and `calories` is still 
positive, albeit a lot weaker. This could mean that `lightly_active_m` has 
less of an impact on calories burned throughout a day.  



### Daily calories  

What's the relationship between calories and day of the week?   

```{r, daily-calories-vs-day-of-week, eval=TRUE}
# Create labels and limits for plot ---------------------------
labels_weekdays <- c(
  "Monday" = "Monday", "Tuesday" = "Tuesday",
  "Wednesday" = "Wednesday", "Thursday" = "Thursday",
  "Friday" = "Friday", "Saturday" = "Saturday",
  "Sunday" = "Sunday"
  )
limits_weekdays <- c(
  "Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
  "Saturday", "Sunday"
  )
# Calories vs day of week plot ---------------------------
ggplot(data = daily_activity) +
  geom_point(
    aes(x = day_of_week, y = calories, color = as.factor(time_of_week)),
    position = "jitter", 
    alpha = 0.3
    ) +
  labs(title = "Daily calories throughout the week", color = "Time of week") +
  ylab("Total Calories") +
  scale_x_discrete(
    "Day",
    labels = labels_weekdays, 
    limits = limits_weekdays,
    guide = guide_axis(angle = 45)
    ) +
  stat_summary(
    aes(x = day_of_week, y = calories),
    fun = mean, 
    geom = "point",
    color = "red", 
    size = 2, 
    alpha = 0.7
    ) +
  theme_minimal()
```

There doesn't look to be a large difference in the average calories burned per 
day. These averages are represented by the red dots on each day.  

### Daily sleep  

What's the relationship between sleep quality and time of week? You might expect 
people to get better sleep over the weekends vs weekdays - are there any 
unexpected trends?  

```{r, sleep-values-vs-time-of-week, eval=TRUE}
# Total minutes asleep vs time of week ---------------------------
ggplot(data = sleep_data) +
  geom_point(
    aes(
    x = weekdays.POSIXt(activity_date), 
    y = total_asleep_merged,
    color = as.factor(time_of_week)
    ),
  position = "jitter",
  alpha = 0.3
  ) +
  labs(title = "Total minutes asleep throughout week") +
  guides(color = "none") +
  ylab("Minutes asleep") +
  scale_x_discrete(
    "Day",
    labels = labels_weekdays, 
    limits = limits_weekdays,
    guide = guide_axis(angle = 45)
    ) +
  stat_summary(
    aes(x = weekdays.POSIXt(activity_date), y = total_asleep_merged),
    fun = mean, 
    geom = "point", 
    color = "red", 
    size = 2,
    alpha = 0.7
    ) +
  theme_minimal()
# Total minutes restless vs time of week ---------------------------
ggplot(data = sleep_data) +
  geom_point(
    aes(
    x = weekdays.POSIXt(activity_date), 
    y = total_restless_merged,
    color = as.factor(time_of_week)
    ),
  position = "jitter",
  alpha = 0.3
  ) +
  labs(title = "Total minutes restless throughout week") +
  guides(color = "none") +
  ylab("Minutes restless") +
  scale_x_discrete(
    "Day",
    labels = labels_weekdays, 
    limits = limits_weekdays,
    guide = guide_axis(angle = 45)
    ) +
  stat_summary(
    aes(x = weekdays.POSIXt(activity_date), y = total_restless_merged),
    fun = mean, 
    geom = "point", 
    color = "red", 
    size = 2,
    alpha = 0.7
    ) +
  theme_minimal()
# Total minutes awake vs time of week ---------------------------
ggplot(data = sleep_data) +
  geom_point(
    aes(
      x = weekdays.POSIXt(activity_date), 
      y = total_awake_merged,
      color = as.factor(time_of_week)
      ),
    position = "jitter",
    alpha = 0.3
    ) +
  labs(title = "Total minutes awake throughout week") +
  guides(color = "none") +
  ylab("Minutes awake") +
  scale_x_discrete(
    "Day",
    labels = labels_weekdays, 
    limits = limits_weekdays,
    guide = guide_axis(angle = 45)
    ) +
  stat_summary(
    aes(x = weekdays.POSIXt(activity_date), y = total_awake_merged),
    fun = mean, 
    geom = "point", 
    color = "red", 
    size = 2,
    alpha = 0.7
    ) +
  theme_minimal()
```

Based on the averages for each day on these graphs (red dots), we can see that:

* The average total minutes sleep were highest on Sunday, Wednesday and 
Saturday 
* The average for minutes awake and minutes restless were highest on Sunday 
and Saturday

People do seem to get sleep more on the weekends. However, it doesn't look like 
the quality of sleep is greater based minutes awake and restless.  


  


# Share  

***

## Stakeholder presentation  

I've provided a HTML file , which includes the following:  

* A summary of my analysis  
* Supporting visualizations and key findings  
* Recommendations for marketing strategy within the Bellabeat App  





